The state learner \\ {-- a super learner for right-censored data}

### 2024-03-21 tex/jrss-version/state-learner-jrss-style.tex
setup with and without figures and table at end

%% Theorem environment
% \theoremstyle{plain} % plain italic style
\newtheorem{theorem}{Theorem}
\numberwithin{theorem}{section}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

@@

@@

\title[The state learner]{The state learner \\ {-- a super learner for right-censored data}}
\author[Anders Munch {\it et al.}]{Anders Munch}
\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
\email{a.munch@sund.ku.dk}
\author{Thomas A.\ Gerds}
\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}

@@

\keywords{Loss based estimation, right-censored data, competing risks, super
  learner, cross-validation}

@@

A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners typically combine strong learners, such as Cox
regression models and random survival forests
\citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.
While the general idea of combining strong learners based on cross-validation
data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
almost as well as the model which minimizes the expected performance, i.e., the
model we would select if we could evaluate the prediction performance in an
infinite hold-out sample.

We are concerned with the choice of the loss function for super learning in
survival analysis. Existing super learner algorithms for right-censored data use
partial log-likelihood loss or inverse probability of censoring weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss restricts the class of learners and
excludes for example simple Kaplan-Meier based learners and also more complex
random survival forest algorithms. For this reason \cite{golmakani2020super}
restrict their learners to Cox proportional hazard models. A lesser known fact
is that a super learner constructed with the negative partial log-likelihood
loss implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. A disadvantage of inverse
probability of censoring weighted loss functions is that they requires a
pre-specified model for the censoring distribution. \cite{westling2021inference}
tackle this challenge by iterating between super learning of the censoring
distribution and the event time distribution.

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously evaluates the loss for learners of the
event time distribution and the censoring distribution. The loss function which
is used to define the state learner is only based on observable
quantities. The state learner can be applied to all types of survival
estimators, works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring distribution. To
analyze the theoretical properties of the state learner we focus on the
so-called discrete super learner which `combines' the library of
learners by picking the one that minimizes the cross-validated loss. The state
learner uses separate libraries to model each competing event and the censoring
distribution. We show that the oracle selector of the state learner is
consistent if all libraries contain a consistent learner and prove a finite
sample oracle inequality.

@@

The article is organized as follows. We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
learning in general with right-censored data, and in
Section~\ref{sec:relat-liter-exist} we review existing methods.
Section~\ref{sec:super-learner-simple} introduces our novel proposal, the state
learner, and Section~\ref{sec:theor-results-prop} provides theoretical
guarantees for the state learner. In Section~\ref{sec:targeted-learning} we
discuss the use of the state learner in the context of targeted learning. We
report some numerical experiments in Section~\ref{sec:numer-exper} and analyze a
prostate cancer data set in Section~\ref{sec:real-data-appl}.
Section~\ref{sec:discussion} contains a discussion of the merits and limitations
of our proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

@@

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). We denote by \(S\) the conditional
event-free survival function,
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M} \) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterized by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the usual right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such that
\(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume that the event
time and the censoring time are conditionally independent given covariates,
\( T \independent C \mid X \). This implies that any distribution
\( P \in \mathcal{P} \) is characterized by a distribution
\( Q \in \mathcal{Q} \) and a conditional cumulative hazard function for \( C \)
given \( X \) \citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard function for
censoring. We assume that \( \Gamma(\blank \mid x) \) is continuous for all
\( x \), and let \(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. In our setting with
competing risks, this yields
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

@@

A super learner estimates a parameter $\Psi$ which can be identified from the
observed data distribution \(P\in\mathcal P\). In this section, to introduce the
concept of super learning, we simply consider estimation of the function
\( \Lambda_{j}\). The parameter \(\Psi:\mathcal P\to\mathcal M\) is then
identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).

As input to the super learner we need a sample \( \data_n=\{O_i\}_{i=1}^n \) of
i.i.d.\ observations from some unknown \( P \in \mathcal{P} \) and a finite
collection of candidate learners $\mathcal{A}$. Each learner
\(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}\) which takes a data set as
input and returns an estimate $a(\data_n) \in \mathcal{M}$ of $\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) using a loss function
\(L\colon \mathcal{M} \times \mathcal{O} \rightarrow \R_+\) by
estimating the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimizes the
average loss according to the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner depend on the
library of learners and on the number of folds \(K\), and that the oracle
learner is a function of the data and the unknown data-generating distribution.
These dependencies are suppressed in the notation.

@@

The problem with most existing super learners for right-censored data is that
they depend on a pre-specified estimator of the censoring distribution. The main
idea of the state learner is to jointly use learners of \( \Lambda_1 \),
\( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners of
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed data. To formally introduce the state learner, we define the
multi-state process
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde D=2\}+ 3\,\1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau].
\end{equation*}
Note that at time \(t\), we observe that each observation is in one of four
mutually exclusive states (Figure \ref{fig:multi-state-process}).
\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{./figure-multi-state-process.pdf}
  \caption{Illustration of the multi-state process \(\eta\) used by
    the state learner. Note that `censored' is a state, hence the
    process is always observed at any time.}
  \label{fig:multi-state-process}
\end{figure}
The conditional distribution of \( \eta(t) \) given \( X=x \) is determined by
the \fxnote*{reconsider letting censoring state be \( -1 \)?}{function}
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
  \quad \text{for all} \quad
  t \in [0,\tau],
  k \in \{0,1,2,3\},
  x \in \mathcal{X}.
\end{equation}
The function \( F \) describes the conditional state occupation
probabilities corresponding to the observed multi-state process
\(\eta\).

@@

$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterizes a \(P\in\mathcal P\) which

@@

In this section we establish theoretical guarantees for the state learner.
% We let \( \E_P \) denote
% expectation under \( P \) and use \( \Theta \) to denote the
% collection of all conditional state-occupation probability functions.
Proposition~\ref{prop:stric-prop} can be derived from the fact that the
integrated Brier score (also called the continuous ranked probability score) is
a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
if we minimize the average loss of the integrated Brier score, we recover the
parameters of the data-generating distribution. In particular, the oracle of a
state learner will be consistent if the library of learners contains at least
one learner that is consistent for estimation of \( F \). Recall that the
function \(F\) implicitly depends on the data generating probability measure
\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
this dependence explicit by writing \(F_0\) for the function which is obtained
by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}.

@@

For all super learners we use a library consisting of three learners: The
Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
model with main effects \citep{cox1972regression, survival-package}, and a
random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
library to learn the outcome and the censoring mechanism. To be precise, note
that given a learner for the cumulative hazard function of the outcome event, we
can typically use the same method to construct a learner of the cumulative
hazard function of the censoring distribution. This would typically work by
training the learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
for the cumulative hazard function of the outcome to learn the cumulative hazard
function of the censoring time, we mean that the learner is trained on
\( \data_n^c \).

>>>>

\usepackage[tablesonly,heads,nolists]{endfloat}

%% Theorem environment
% \theoremstyle{plain} % plain italic style
\newtheorem{theorem}{Theorem}
% \numberwithin{theorem}{section}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}

@@

%% Toggle with or without figures
\def\nofig{1}

%% TODO: Update figures with name 'figure-'.
%% Write macro for including figure or not.

% \newcommand{\includeFigCond}[2]{
%   \ifx\nofig\undefined %
%     \includegraphics[#1]{#2} %
%   \else %
%     \texttt{#2} %
%   \fi %
% }
\newcommand{\includeFigCond}[2][]{
  \ifx\nofig\undefined %
    \includegraphics[#1]{#2} %
  \else %
    \texttt{#2} %
  \fi %
}

@@

\title[The state learner]{The state learner \\ {-- a super learner for right-censored data}}
\author{Anders Munch}
\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}
\email{a.munch@sund.ku.dk}
\author[Munch {\it et al.}]{Thomas A.\ Gerds}
\address{Section of Biostatistics, University of Copenhagen, Copenhagen, Denmark}

@@

\keywords{Competing risks, cross-validation, loss based estimation,
  right-censored data, super learner}

@@

A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners typically combine strong learners, such as Cox
regression models and random survival forests
\citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.
While the general idea of combining strong learners based on cross-validation
data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
almost as well as the model which minimises the expected performance, i.e., the
model we would select if we could evaluate the prediction performance in an
infinite hold-out sample.

We are concerned with the choice of the loss function for super learning in
survival analysis. Existing super learner algorithms for right-censored data use
partial log-likelihood loss or inverse probability of censoring weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss restricts the class of learners and
excludes for example simple Kaplan-Meier based learners and also more complex
random survival forest algorithms. For this reason \cite{golmakani2020super}
restrict their learners to Cox proportional hazard models. A lesser known fact
is that a super learner constructed with the negative partial log-likelihood
loss implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. A disadvantage of inverse
probability of censoring weighted loss functions is that they require a
pre-specified model for the censoring distribution. \cite{westling2021inference}
tackle this challenge by iterating between super learning of the censoring
distribution and the event time distribution.

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously evaluates the loss for learners of the
event time distribution and the censoring distribution. The loss function which
is used to define the state learner is only based on observable
quantities. The state learner can be applied to all types of survival
estimators, works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring distribution. To
analyze the theoretical properties of the state learner we focus on the
so-called discrete super learner which `combines' the library of
learners by picking the one that minimises the cross-validated loss. The state
learner uses separate libraries to model each competing event and the censoring
distribution. We show that the oracle selector of the state learner is
consistent if all libraries contain a consistent learner and prove a finite
sample oracle inequality.

@@

The article is organised as follows. We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
learning in general with right-censored data, and in
Section~\ref{sec:relat-liter-exist} we review existing methods.
Section~\ref{sec:super-learner-simple} introduces our novel proposal, the state
learner, and Section~\ref{sec:theor-results-prop} provides theoretical
guarantees for the state learner. In Section~\ref{sec:targeted-learning} we
discuss the use of the state learner in the context of targeted learning. We
report some numerical experiments in Section~\ref{sec:numer-exper} and analyze a
prostate cancer data set in Section~\ref{sec:real-data-appl}.
Section~\ref{sec:discussion} contains a discussion of the merits and limitations
of our proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

@@

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). We denote by \(S\) the conditional
event-free survival function,
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M} \) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the usual right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such that
\(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume that the event
time and the censoring time are conditionally independent given covariates,
\( T \independent C \mid X \). This implies that any distribution
\( P \in \mathcal{P} \) is characterised by a distribution
\( Q \in \mathcal{Q} \) and a conditional cumulative hazard function for \( C \)
given \( X \) \citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard function for
censoring. We assume that \( \Gamma(\blank \mid x) \) is continuous for all
\( x \), and let \(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. In our setting with
competing risks, this yields
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

@@

A super learner estimates a parameter $\Psi$ which can be identified from the
observed data distribution \(P\in\mathcal P\). In this section, to introduce the
concept of super learning, we simply consider estimation of the function
\( \Lambda_{j}\). In this case, the parameter \(\Psi:\mathcal P\to\mathcal M\)
is identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).

As input to the super learner we need a sample \( \data_n=\{O_i\}_{i=1}^n \) of
i.i.d.\ observations from some unknown \( P \in \mathcal{P} \) and a finite
collection of candidate learners $\mathcal{A}$. Each learner
\(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}\) which takes a data set as
input and returns an estimate $a(\data_n) \in \mathcal{M}$ of $\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) using a loss function
\(L\colon \mathcal{M} \times \mathcal{O} \rightarrow \R_+\) by
estimating the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimises the
average loss according to the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner depend on the
library of learners and on the number of folds \(K\), and that the oracle
learner is a function of the data and the unknown data-generating distribution.
These dependencies are suppressed in the notation.

@@

The problem with most existing super learners for right-censored data is that
they depend on a pre-specified estimator of the censoring distribution. The main
idea of the state learner is to jointly use learners of \( \Lambda_1 \),
\( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed data
distribution \( P \). The discrete state learner ranks a tuple of learners of
\( (\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model the
observed data. To formally introduce the state learner, we define the
multi-state process
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde D=2\}+ 3\,\1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau].
\end{equation*}
Note that at time \(t\), we observe that each observation is in one of four
mutually exclusive states (Figure \ref{fig:multi-state-process}).
\begin{figure}[h]
  \centering %
  \includeFigCond[width=.5\textwidth]{figure-multi-state-process.pdf}
  \caption{Illustration of the multi-state process \(\eta\) used by
    the state learner. Note that `censored' is a state, hence the
    process is always observed at any time.}
  \label{fig:multi-state-process}
\end{figure}
The conditional distribution of \( \eta(t) \) given \( X=x \) is determined by
the \fxnote*{reconsider letting censoring state be \( -1 \)?}{function}
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
  \quad \text{for all} \quad
  t \in [0,\tau],
  k \in \{0,1,2,3\},
  x \in \mathcal{X}.
\end{equation}
The function \( F \) describes the conditional state occupation
probabilities corresponding to the observed multi-state process
\(\eta\).

@@

$(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a \(P\in\mathcal P\) which

@@

In this section we establish theoretical guarantees for the state learner.
% We let \( \E_P \) denote
% expectation under \( P \) and use \( \Theta \) to denote the
% collection of all conditional state-occupation probability functions.
Proposition~\ref{prop:stric-prop} can be derived from the fact that the
integrated Brier score (also called the continuous ranked probability score) is
a strictly proper scoring rule \citep{gneiting2007strictly}. This implies that
if we minimise the average loss of the integrated Brier score, we recover the
parameters of the data-generating distribution. In particular, the oracle of a
state learner will be consistent if the library of learners contains at least
one learner that is consistent for estimation of \( F \). Recall that the
function \(F\) implicitly depends on the data generating probability measure
\(P\in\mathcal P\) but that this was suppressed in the notation. We now make
this dependence explicit by writing \(F_0\) for the function which is obtained
by substituting a specific \(P_0\in\mathcal{P}\) for \(P\) in equation
\eqref{eq:transition}.

@@

For all super learners we use a library consisting of three learners: The
Kaplan-Meier estimator \citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox
model with main effects \citep{cox1972regression, survival-package}, and a
random survival forest \citep{ishwaran2008random,rfsrc-paclage}. We use the same
library to learn the outcome and the censoring mechanism. To be precise, note
that given a learner for the cumulative hazard function of the outcome event, we
can typically use the same method to construct a learner of the cumulative
hazard function of the censoring distribution. This works by training the
learner on the data set \( \data_n^c \), where
\( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta, X_i) \). When we say that we use a learner
for the cumulative hazard function of the outcome to learn the cumulative hazard
function of the censoring time, we mean that the learner is trained on
\( \data_n^c \).
